Automatic Matching of Bullet Lands

### 2016-04-15 imaging-paper.Rnw
A couple small changes
A few other small changes

\address{Address of the Authors\\
Department of Statistics and Statistical Laboratory\\ Snedecor Hall\\
Ames Iowa 50011-1210\\
\printead{e1}\\
\phantom{E-mail:\ }\printead*{e2}\\
\phantom{E-mail:\ }\printead*{e3}}
\end{aug}
%
\begin{abstract}
In 2009, the National Academy of Sciences published a report questioning the scientific validity of many forensic methods including firearm examination. Firearm examination is a forensic tool used to help the court determine whether two bullets were fired from the same gun barrel. During the firing process, rifling, manufacturing defects, and impurities in the barrel create striation marks on the bullet. Identifying these striation markings in an attempt to match two bullets is the primary goal. We propose an automated framework for the analysis of the 3D surface measurements, that first transcribes the markings into a 2D plotting framework. This makes identification of matches easier and allows for a quantification of both matches and matchability of barrels. The automatic matching routine we propose manages to (a) correctly identify lands with too much damage to be suitable for matching, and (b) correctly identify all 10,384 land-to-land matches of the James Hamby study \citep{hamby:2009}.
\end{abstract}

@@

Throughout this paper, we work with images from the James Hamby Consecutively Rifled Ruger Barrel Study~\citep{hamby:2009}. Ten consecutively rifled Ruger P-85 pistol barrels were obtained from the manufacturer and fired to produce 20 known test bullets and 15 unknown bullets for comparison. 
3D topographical images of each bullet were obtained using a NanoFocus lens at 20x magnification and made publicly available on the NIST Ballistics Database Project\footnote{\url{http://www.nist.gov/forensics/ballisticsdb/hamby-consecutively-rifled-barrels.cfm}} in a format called x3p (XML 3-D Surface Profile). The x3p format conforms to the ISO5436-2 standard\footnote{\url{http://sourceforge.net/p/open-gps/mwiki/X3p/}}, implemented to provide a simple and standard conforming way to exchange 2D and 3D profile data. It was adopted by the OpenFMC (Open Forensic Metrology Consortium\footnote{\url{http://www.openfmc.org/}}), a group of academic, industry, and government firearm forensics researchers whose aim is to establish best practices for researchers using metrology in the forensic sciences.

@@

\subsection{Horizontal alignment}
Signatures of each of the two lands, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in this representations as peaks and valleys.  The individual characteristics are quite prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lof$bulletname <- scrubPath(lof$bullet)
qplot(y, resid, group=bulletname, colour=bulletname, data=lof, 
      geom="line") + 
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}    
\begin{subfigure}[b]{.49\textwidth}\centering
    \caption{Aligned signatures.\label{fig:crosscutX}}{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lofY$bulletname <- scrubPath(lofY$bullet)
qplot(y+min(lof$y), resid, group=bulletname, colour=bulletname, data=lofY, 
      geom="line") + xlab("y") +
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}
\caption{\label{fig:cross100}Signatures of bullets 1-5 and 2-1 taken at  heights of $x = 100\mu m$. A horizontal shift of the values of bullet~1-5 to the right shows the similarity of the striation marks.}
\end{figure}
%
This horizontal shift is based on the cross-correlation between the two signatures: let $f(t)$ and $g(t)$ define the signature values  at $t$, where $t$ are locations between 0~$\mu m$ and about 2500~$\mu m$, 1.5625~$\mu m$ apart.
The cross-correlation between $f$ and $g$ at lag $k$ is then defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).

\begin{figure}[hbtp]
  \centering
<<ccf, echo=FALSE, fig.width = 8, fig.height = 3, out.width = '.65\\textwidth'>>=
lag <- ccf$lag[which.max(ccf$acf)]
qplot(x=ccf$lag, xend=ccf$lag, y = 0, yend= ccf$acf, geom="segment", colour = I("grey50")) + 
  theme_bw() + ylab("Correlation") + xlab("Lag k") +
  geom_segment(aes(x = lag, xend = lag, y = 0, yend =  max(ccf$acf)), colour="black") + 
  scale_x_continuous(breaks = c(-100, -50, lag, 0, 50, 100),
                     minor_breaks = c(-75,-25,25, 75))
@
\caption{\label{fig:ccf}Cross-correlation function between the two signatures shown in Figure~\ref{fig:crosscut} at lags between -100 and 100. At a lag of -17 the correlation peaks, indicating the largest amount of agreement between the signatures. Figure~\ref{fig:crosscutX} shows the lag-shifted signatures.}
\end{figure}

>>>>

\address{Address of the Authors\\
Department of Statistics and Statistical Laboratory\\ Snedecor Hall\\
Ames Iowa 50011-1210\\
\printead{e1}\\
\phantom{E-mail:\ }\printead*{e2}\\
\phantom{E-mail:\ }\printead*{e3}}
\end{aug}
%
\begin{abstract}
In 2009, the National Academy of Sciences published a report questioning the scientific validity of many forensic methods including firearm examination. Firearm examination is a forensic tool used to help the court determine whether two bullets were fired from the same gun barrel. During the firing process, rifling, manufacturing defects, and impurities in the barrel create striation marks on the bullet. Identifying these striation markings in an attempt to match two bullets is the primary goal. We propose an automated framework for the analysis of the 3D surface measurements, that first transcribes the markings into a 2D plotting framework. This makes identification of matches easier and allows for a quantification of both matches and matchability of barrels. The automatic matching routine we propose manages to (a) correctly identify lands with too much damage to be suitable for comparison, and (b) correctly identify all 10,384 land-to-land matches of the James Hamby study \citep{hamby:2009}.
\end{abstract}

@@

Throughout this paper, we work with images from the James Hamby Consecutively Rifled Ruger Barrel Study~\citep{hamby:2009}. Ten consecutively rifled Ruger P-85 pistol barrels were obtained from the manufacturer and fired to produce 20 known test bullets and 15 unknown bullets for comparison. 
3D topographical images of each bullet were obtained using a NanoFocus lens at 20x magnification and made publicly available on the NIST Ballistics Database Project\footnote{\url{http://www.nist.gov/forensics/ballisticsdb/hamby-consecutively-rifled-barrels.cfm}} in a format called x3p (XML 3-D Surface Profile). The x3p format conforms to the ISO5436-2 standard\footnote{\url{http://sourceforge.net/p/open-gps/mwiki/X3p/}}, implemented to provide a simple and standard conforming way to exchange 2D and 3D profile data. It was adopted by the OpenFMC (Open Forensic Metrology Consortium\footnote{\url{http://www.openfmc.org/}}), a group of academic, industry, and government firearm forensics researchers whose aim is to establish best practices for researchers using metrology in forensic science.

@@

\subsection{Horizontal alignment}
Signatures of each of the two lands, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in these representations as peaks and valleys.  The individual characteristics are quite prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lof$bulletname <- scrubPath(lof$bullet)
qplot(y, resid, group=bulletname, colour=bulletname, data=lof, 
      geom="line") + 
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}    
\begin{subfigure}[b]{.49\textwidth}\centering
    \caption{Aligned signatures.\label{fig:crosscutX}}{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lofY$bulletname <- scrubPath(lofY$bullet)
qplot(y+min(lof$y), resid, group=bulletname, colour=bulletname, data=lofY, 
      geom="line") + xlab("y") +
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}
\caption{\label{fig:cross100}Signatures of bullets 1-5 and 2-1 taken at  heights of $x = 100\mu m$. A horizontal shift of the values of bullet~1-5 to the right shows the similarity of the striation marks.}
\end{figure}
%
This horizontal shift is based on the cross-correlation between the two signatures: let $f(t)$ and $g(t)$ define the signature values  at $t$, where $t$ are locations between 0~$\mu m$ and about 2500~$\mu m$, 1.5625~$\mu m$ apart.
The cross-correlation between $f$ and $g$ at lag $k$ is then defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).

\begin{figure}[hbtp]
  \centering
<<ccf, echo=FALSE, fig.width = 8, fig.height = 3, out.width = '.65\\textwidth'>>=
lag <- ccf$lag[which.max(ccf$acf)]
qplot(x=ccf$lag, xend=ccf$lag, y = 0, yend= ccf$acf, geom="segment", colour = I("grey50")) + 
  theme_bw() + ylab("Correlation") + xlab("Lag k") +
  geom_segment(aes(x = lag, xend = lag, y = 0, yend =  max(ccf$acf)), colour="black") + 
  scale_x_continuous(breaks = c(-100, -50, lag, 0, 50, 100),
                     minor_breaks = c(-75,-25,25, 75))
@
\caption{\label{fig:ccf}Cross-correlation function between the two signatures shown in Figure~\ref{fig:crosscut} at lags between -100 and 100. The correlation peaks at a lag of -17, which indicates the largest amount of agreement between the signatures. Figure~\ref{fig:crosscutX} shows the lag-shifted signatures.}
\end{figure}
